Co-methylation

### 2015-04-22 Rmarkdown/comethylation.Rmd
Massive improvements to simulation model exploring different estimators of the odds ratio.

```{r}
# Function to simulate K 2-tuples with Poisson(mean_d) sequencing coverage and marginal
# probabilities p1, p2
simTables <- function(K, mean_d, p1, p2) {
  .simTables <- function(mean_d, p1, p2) {

    d <- rpois(1, mean_d)
    z1 <- factor(rbinom(d, 1, p1), levels = c(0, 1))
    z2 <- factor(rbinom(d, 1, p2), levels = c(0, 1))
    xtabs(~ z2 + z1, drop.unused.levels = FALSE)
  }
  replicate(K, .simTables(mean_d, p1, p2))
}

# Estimator of the (offset-adjusted) unconditional log odds ratios of K 2x2 tables
LOR <- function(x, offset = 0.5) {
  MM <- x[2, 2, ] + offset
  MU <- x[2, 1, ] + offset
  UM <- x[1, 2, ] + offset
  UU <- x[1, 1, ] + offset
  log(MM) + log(UU) - log(MU) - log(UM)
}

# Mantel-Haenszel estimator of a common log odds ratio of a 2x2xK table
MH <- function(x, offset = 0.5) {
  # TODO: return the standard error
  MM <- x[2, 2, ]
  MU <- x[2, 1, ]
  UM <- x[1, 2, ]
  UU <- x[1, 1, ]
  d <- apply(x, 3, sum)
  or_mh <- sum((UU * MM) / d) / sum((MU * UM) / d)
  log(or_mh)
}

# Estimator of the (offset-adjusted) unconditional log odds ratio of the
# 2x2xK collapsed over K.
LORCollapsed <- function(x, offset) {
  z <- apply(x, c(1, 2), sum)
  # Artificially augment so it's a 2x2x1 table
  dim(z) <- c(dim(z), 1)
  LOR(z)
}

# Simulate and plot data
simAndPlot <- function(K, mean_d, p1, p2) {
  x <- mapply(function(p1, p2, K, mean_d) {
    simTables(K = K, mean_d = mean_d, p1 = p1, p2 = p2)
  }, p1 = p1, p2 = p2, MoreArgs = list(K = K, mean_d = mean_d), SIMPLIFY = FALSE)

  par(mfrow = c(3, 3))
  mapply(function(x, p1, p2) {
    hist(LOR(x), main = paste0("p1 = ", p1, ", p2 = ", p2), xlim = c(-8, 8), breaks = 16)
    abline(v = MH(x), col = "dodgerBlue", lwd = 3)
    abline(v = LORCollapsed(x), col = "orange", lwd = 3)
    abline(v = 0, col = "black", lty = 2)
  }, x = x, p1 = p1, p2 = p2)
  title(paste0("K = ", K, ", mean_d = ", mean_d), outer = TRUE, line = -1)
}

# Vary p1 with p2 = p1 at 30x coverage across 1000 CpG pairs
K <- 1000
mean_d <- 30
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
simAndPlot(K, mean_d, p1, p2)

# Vary p1 with p2 = 1 - p1 at 30x coverage across 1000 CpG pairs
K <- 1000
mean_d <- 30
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- 1 - p1
simAndPlot(K, mean_d, p1, p2)

# Vary p1 with p2 = p1 at 1000x coverage across 1000 CpG pairs
K <- 1000
mean_d <- 1000
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
simAndPlot(K, mean_d, p1, p2)

# TODO: p1 and p2 distributed U(0, 1) at 30x coverage across 1000 CpG pairs
# Will need to simulate p1 and p2 within .simTable().
```

@@

>>>>

__TODO: Move this code chunk to supplementary material and only include results in thesis.__

```{r}
# Requires the installation of methsim
library(mipfp)

# rtpois is a helper function to simulate n truncated Poisson(lambda) with
# truncation specified by 'trunc'. 
# It is adapted from a method for simulating zero-truncated Poisson suggested 
# by Peter Dalgaard https://stat.ethz.ch/pipermail/r-help/2005-May/070680.html
rtpois <- function(n, lambda, trunc) {
  qpois(runif(n, ppois(trunc, lambda), 1), lambda) 
}

# Function to simulate 2x2xK tables.
# Each 2x2 table represents the counts of unmethylated (0) and methylated (1) states
# at each CpG for each read.
# The number of reads per table (i.e., 2x2 table sample size) is simulated 
# from a truncated Poisson(lambda) distribution; truncation is to simulate
# minimum coverage cutoffs and is specified by the 'trunc' argument.
# The marginal probabilities of each table are p1 = Prob(methylated at first CpG)
# and p2 = Prob(methylated at second CpG).
# The odds ratio (or) is for each 2x2 table and should be of length K or 1 (in
# which case all K tables have the same odds ratio).
simTables <- function(K, lambda, trunc, p1, p2, or) {

  # d is the sample size of each 2x2 table (i.e, the sequencing depth for the 2-tuple)
  d <- rtpois(n = K, lambda = lambda, trunc = trunc)
  if (length(p1) == 1L) {
    p1 <- rep(p1, K)
  }
  if (length(p2) == 1L) {
    p2 <- rep(p2, K)
  }
  if (length(or) == 1L) {
    or <- rep(or, K)
  }
  .simTables <- function(k, d, p1, p2, or) {
    # Compute joint probabilities
    P <- mipfp::Ipfp(seed = matrix(c(or, 1, 1, 1), ncol = 2), 
                     target.list = list(1, 2), 
                     target.data = list(c(1 - p1, p1), c(1 - p2, p2)))$x.hat
    matrix(rmultinom(1, d, as.vector(P)), ncol = 2)
  }
  mapply(.simTables, k = K, d = d, p1 = p1, p2 = p2, or = or, SIMPLIFY = "array")
}

# Estimator of the (offset-adjusted) unconditional log odds ratios of K 2x2 tables
LOR <- function(x, offset = 0.5) {
  x <- x + offset
  MM <- x[2, 2, ]
  MU <- x[2, 1, ]
  UM <- x[1, 2, ]
  UU <- x[1, 1, ]
  log2(MM) + log2(UU) - log2(MU) - log2(UM)
}

# Mantel-Haenszel estimator of a common log odds ratio of a 2x2xK table
MH <- function(x, offset = 0) {
  x <- x + offset
  # TODO: return the standard error
  MM <- x[2, 2, ] 
  MU <- x[2, 1, ]
  UM <- x[1, 2, ]
  UU <- x[1, 1, ]
  d <- apply(x, 3, sum)
  or <- sum((UU * MM) / d) / sum((MU * UM) / d)
  # One or more margins must be identically zero so add offset
  if (is.nan(or) | isTRUE(all.equal(or, 0)) | is.infinite(or)) {
    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
    MH(x, 0.5)
  } else {
    log2(or)
  }
}

# Estimator of the (offset-adjusted) unconditional log odds ratio of the
# 2x2xK collapsed over K.
LORCollapsed <- function(x, offset = 0.5) {
  z <- apply(x, c(1, 2), sum)
  # Artificially augment so it's a 2x2x1 table
  dim(z) <- c(dim(z), 1)
  lor <- LOR(z, 0)
  if (is.nan(lor) | is.infinite(lor)) {
    warning("One or more cell is (probably) identically zero in all K 2x2 tables. Adding offset = 0.5")
    LOR(z, 0.5)
  } else {
    lor
  }
}

# Simulate and plot data when p1 = p2.
# 100 CpG pairs, 30x average coverage truncated at 10
# odds ratio = 1, i.e., independence
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 1
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i], p2 = p2[i], or = or)
  hist(LOR(x), main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)), 
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 = p2.
# 1000 CpG pairs, 30x average coverage truncated at 10
# odds ratio = 1, i.e., independence
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 10
or <- 1
p1 <- c(0.01, 0.05, 0.1, 0.25, 0.5, 0.75, 0.9, 0.95, 0.99)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1[i], 
                 p2 = p2[i], or = or)
  hist(LOR(x), 
       main = paste0("p1 = ", p1[i], ", p2 = ", p2[i], ", lor = ", log2(or)), 
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1, p2 ~ Uniform(0, 1)
# 100 CpG pairs, 30x average coverage truncated at 10
# log odds ratio = -4, -3, ..., 3, 4.
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 2^seq(-4, 4, 1)
p1 <- runif(K)
p2 <- runif(K)
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, 
                 or = or[i])
  hist(LOR(x), 
       main = paste0("p1 = U(0, 1), p2 = U(0, 1)", ", lor = ", log2(or[i])),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or[i]), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), 
      outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Uniform(0, 1), p2 = p1
# 100 CpG pairs, 30x average coverage truncated at 10
# log odds ratio = -4, -3, ..., 3, 4.
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 2^seq(-4, 4, 1)
p1 <- runif(K)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
  hist(LOR(x), main = paste0("p1 = U(0, 1), p2 = p1", ", lor = ", log2(or[i])),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       frequeny = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or[i]), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), 
      outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 100 CpG pairs, 30x average coverage truncated at 10
# log odds ratio = -4, -3, ..., 3, 4.
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or <- 2^seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or[i])
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = ", log2(or[i])),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  abline(v = log2(or[i]), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 100 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ Gaussian(log2(or_mean), 1)
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 100
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ Gaussian(log2(or_mean), 1)
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = Norm(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ Skewed-Gaussian(log2(or_mean), 1, -10)
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
# levels (B).
library(sn)
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rsn(K, log2(or_mean[i]), 1, -10)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = SN(", log2(or_mean[i]), ", 1, -10)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)

# Simulate and plot data when p1 ~ Beta(0.3, 0.2), p2 = p1.
# 1000 CpG pairs, 30x average coverage truncated at 10
# log2(or) ~ 0.5 * Gaussian(log2(or_mean), 1) + 0.5 * Gaussian(-log2(or_mean))
# Beta(0.3, 0.2) is roughly like the genome-wide distribution of methylation 
# levels (B).
set.seed(666)
par(mfrow = c(3, 3))
K <- 1000
lambda <- 20
trunc <- 11
or_mean <- 2 ^ seq(-4, 4, 1)
w <- sample(c(-1, 1), K, replace = TRUE)
p1 <- rbeta(K, 0.3, 0.2)
p2 <- p1
for (i in 1:9) {
  or <- 2 ^ rnorm(K, w * log2(or_mean[i]), 1)
  x <- simTables(K = K, lambda = lambda, trunc = trunc, p1 = p1, p2 = p2, or = or)
  hist(LOR(x), main = paste0("p1 = Beta(0.3, 0.2), p2 = p1", ", lor = NM(", log2(or_mean[i]), ", 1)"),
       xlim = c(-10, 10), breaks = 20,
       xlab = expression(paste(log[2], "(odds ratio)")),
       freq = FALSE)
  abline(v = MH(x), col = "dodgerBlue", lwd = 3)
  abline(v = LORCollapsed(x), col = "orange", lwd = 3)
  lines(density(log2(or)), col = "black", lty = 2)
}
title(paste0("K = ", K, ", lambda = ", lambda, ", trunc = ", trunc), outer = TRUE, line = -1)
```

@@

## Plan of attack

Following discussion with Terry (2015-04-21), this is my plan of attack to finish this chapter and the `methsim` chapter.

- [x] Finish reading Greenland (1982).
  - What happens with the MH estimator when the common odds ratio assumption is invalid. Is there a better estimator in this setting?

  > As a consequence, when there is heterogeneity, the expectation of MH varies with the overall and stratum-specific case-control ratios. The usual common odds ratio estimators share this disconcerting property \citep{Greenland:1982uk}.

  In contrast to the MH estimator, the Miettenen estimator consistently estimates meaningful epimiological parameters. However, it "[requires] consideration in the selection of the weights".

- [ ] Demonstrate via simulation that the unconditional LOR estimator applied to $2 \times 2$ tables is biased when $d$ is small and the marginal probabilities are near the boundaries.
  - Corrollary: Don't use the LOR estimates of within-fragment co-methylation from individual $2\times 2$ tables. I know they're biased, so there's no reason to show these results.
- [ ] Demonstrate that the MH estimator applied to $2 \times 2 \times K$ tables is unbiased when $d$ is small (per-table) and the marginal probabilities are near the boundaries.
  - Need to investigate the setting where the marginal probabilities are heterogeneous across the $K$ tables but with the same odds-ratio.
  - Need to investigate the setting where the odds ratio __is not__ common to the $K$ tables; what does the MH estimator end up estimating?

- Run Woolf's test or Breslow-Day test of homogeneity of odds ratios (noting that these are not powerful tests).

__In fact, Breslow himself advocates for the Tarone statistic in place of the Breslow-Day statistic (Breslow, N. E. JASA 91, 14–28 (1996)). Also, Woolf's test, Breslow-Day and Tarone's statistic are only appropriate in Asymptopia I (a small number of tables with large frequencies) and not in Asymptopia II (a large number of tables with small frequencies); see Liang and Self (1985) for Asymptopia II (the scenario I am in)__

Liang, K.-Y. & Self, S. G. Biometrika (1985) advocate for a conditional test of homogeneity of odds ratios when the data are sparse (Asymptopia II) but note that:

> The problem with the conditional approach is its complexity. It is clear that when the sample size of each table increases, the computation of T2, T4 or T5 [ the conditional tests] becomes a burden which, eventually, will be prohibitive.
This raises the question of how large the sample sizes in each table must be before the unconditional score test T1 is reasonable to use. Further simulation results suggest that even when the number of cases and controls are as small as 10 with as many as 10 tables, the large-sample chi-squared approximation is adequate in that the observed sizes are near to the nominal 0 05 level.

I have $K \\g 10$ ($K$ = number of $2 \times 2$ tables) and $d \geq 10$ ($d$ = sequencing coverage), so the unconditional test may well be fine.

__TODO: Look up Liang-Self in Agresti's CDA (2013)__

- Algorithm: If the null of homogeneity of odds ratio is rejected => stratify further(?), otherwise compute MH or Miettenen estimates of the log odds ratio (and standard errors) for different stratification strategies (see [https://stat.ethz.ch/pipermail/r-help/2007-February/126254.html](https://stat.ethz.ch/pipermail/r-help/2007-February/126254.html))
  - Stratify by IPD
  - Stratify by IPD + chromosome
  - Stratify by IPD + chromosome + chromosome arm
  - Stratify by IPD + chromosome + chromosome arm + bin (Mb)
  - Check whether the more-stratified estimates fall within a reasonable distance from the less-stratified estimates (i.e., with respect to standard errors of the less-stratified estimates).
- Settle on the most appropriate level of stratification and apply to all samples.
  - Discuss results.
- `methsim`: Sample LORs from a Normal(mu, sigma^2) distribution, where mu (resp. sigma) is the MH estimator (resp. standard error? empirical variance of stratified MH estimators?) of within-fragment co-methylation for that IPD and region-type.
